\chapter{Petri Nets}
Petri Nets is a modeling languages notation that allow us to apply different analysis approaches. Petri nets have been proposed to model concurrent systems. 

\section{Concurrent Systems}
A concurrent system is a system (not strictly from computer science) where there are processes of any type executed at any time. In this system you have a workflow where some processes works in parallel (independent from each other) and others work sequentially (they must wait for the previous machines to finish).\par

Petri Nets are a graphical model notation for concurrent systems in which you construct a graph that describe resources and events involving those resources. There are two types of nodes:

\begin{itemize}
    \item \textbf{Places:} represent type of resources. 
        \begin{itemize}
            \item \textbf{Tokens:} represent instances of a resource of some specific type. Tokes stay inside places. The place where they are describe their type
        \end{itemize}
    \item \textbf{Transition:} represent events that change the state of one or more resources. They connect places to places (like a Multiset Rewriting rule or a chemical reaction). Incoming edges are connected to places that provide resources (tokens) that are consumed by the transition, while output edges represent places for which resources have been produced.
\end{itemize}

\section{Firing a transition}
Firing a transition means \textbf{executing it}. In order to fire a transition you must have enough resources in input, then it will produce the necessary resources for the output places.

\section{Formal definition of Petri Nets}
A Petri Net is a tuple $<P, T>$ where:
\begin{itemize}
    \item $P$ is the (finite) set of \textbf{places}.
    \item $T$ is the (finite) set of \textbf{transitions} s.t. each transition $t$ is a tuple $<I, O>$ where:
        \begin{itemize}
            \item $I$ is a function s.t. $t$ consumes $I(p)$ tokens in each place $p$.
            \item $O$ is a function s.t. $t$ produces $O(p)$ tokens in each place $p$.
        \end{itemize}
\end{itemize}

\subsection{Markings}
Tokens that are placed in the network represent\textbf{ the state of the network}. The state of the Petri Net is also called the \textbf{marking of the Petri Net}. A marking con be seen in two different ways:

\begin{itemize}
    \item a function $m$ s.t. $m(p)$ is the number of tokens in place $p$.
    \item a vector $m = <m_{1}, m_{2}, ..., m_{n}$ where $m_{i}$ is the number of tokens in place $p_{i}$. 
    \item a vector $m = <d_{1}p_{1}, ..., d_{n}p_{n}$ where $d_{i}$ represents the number of tokens in $p_{i}$.
\end{itemize}

\subsection{Firing a transition}
Given a transition $t = <I, O>$, it can be fired from $m$ $iff$ for any place p:

\begin{equation*}
    m(p) \geq l(p)
\end{equation*}

If the condition is satisfied then the firing transforms the marking $m$ into a marking $m^{'}$ s.t. for any place p:
\begin{equation*}
    m^{'}(p) = m(p) - I(p) + O(p)
\end{equation*}

This transformation can be described by the following notations:
\begin{itemize}
    \item $m \rightarrow m^{'}$
    \item $Post(m) = \{m^{'} | m \rightarrow m^{'}\}$. $Post(m)$ is the set of markings that can be obtained by firing transitions from $m$ ($Post$ correspond to the transition relation in the Transition Systems)
\end{itemize}

\subsubsection{Initial marking}
The marking that we consider to be the first one is colled the \textbf{initial marking} $m_{0}$.

\subsubsection{Reachable markings}
Given two markings $m$ and $m{'}$ s.t.:
\begin{equation*}
    m \rightarrow m_{1} \rightarrow m_{2} \rightarrow ... \rightarrow m^{'}
\end{equation*}
Then we can say that $m^{'}$ is reachable from $m$.\par
Given the Petri Net $N$ with initial marking $m_{0}$, then the function $Reach(N)$ is the set of reachable markings of N s.t.:
\begin{center}
    Reach(N) = \{$m$ reachable from $m_{0}$\}
\end{center}

\subsection{Ordering on markings}
Markings can be compared using the precedence relation $\preceq$:
\begin{itemize}
    \item $m \preceq m^{'} \iff \forall p. m(p) \leq m^{'}(p)$.
    \item $m \prec m^{'} \iff m \preceq m^{'} \land m \neq m^{'}$
\end{itemize}

\section{What can we use Petri Nets for}
The question we can investigate on Petri Nets are these:
\begin{itemize}
    \item \textbf{Boundedness:} is the number of reachable markings bounded?
    \item \textbf{Place boundedness:} in each place the number of tokens I can obtain is bounded or can it grows in an unbounded way?
    \item \textbf{Semi-liveness:} the transition that are present in a Petri Net will they all be fired? There are some transition that can never be fired?
    \item \textbf{Coverability}
\end{itemize}

\section{Decidability of a marking}
The reachability graph of a Petri Net (aka the Transition System of a MultiSet Rewriting system) can be infinite, \textbf{but the reachability property is decidable}. The computation of decidability has been proven to require \textbf{exponential time}. This means that Petri Net are not Turing equivalent.\par
In order to compute reachability in an efficient way, several alternatived have been proposed:

\subsubsection{Solution 1}
We consider overapproximations of the set of reachable states (based on Place Invariants or on Karp and Miller tree). We compute in polynomial time a set which is an overapproximation of the set of reachable marking. So our solution will include surely ALL the reachable marking, but also it can contain some unreachable ones. So we are only sure that if a marking is not in the overapproximation, then it is not reachable.

\subsubsection{Solution 2}
Instead of consider reachability, we will answer coverability: "is a marking greater or equal than is reachable?" The problem of coverability is weaker than reachability, but easy to compute and meaningful in the context of Petri Nets.

\section{Place Invariants}
Place Invariants are a way to reason about marking which has some analogy which the mass conservation invariant (matter can never be created or destroyed). This can be used to put constraint on a set of reachable markings.\par
Formally a Place invariant (or p-semiflow) is a vector $i$ of natural numbers s.t. for any reachable marking m:
\begin{equation*}
    \sum_{p \in P} i(p) \times m(p) = \sum_{p \in P} i(p) \times m_{0}(p)
\end{equation*}

\subsection{Invariants as overapproximations}
Given a marking $m$ and an invariant $i$, we can say that if $m$ is reachable and $i$ is an invariant, then:
\begin{equation*}
    \sum_{p \in P} i(p) \times m(p) = \sum_{p \in P} i(p) \times m_{0}(p)
\end{equation*}
The reverse is not true.

\subsubsection{Theorem}
$\forall$ Petri Net N:
\begin{center}
    $Reach (N) \subseteq \{ m | m $ respects some invariant of $N\}$
\end{center}

So if a marking does not respect the invariant, \textbf{then we surely know that is not reachable}.

\subsubsection{Theorem: place invariant and boundedness}
If there exists a place invariant $i$ and a place $p$ s.t. $i(p) > 0$, then $p$ is bounded. \textbf{The reverse is not true.}

\section{How to compute Place invariants}
\subsection{Matrix chatacterisation}
Let us represent the Petri Net as a matrix where we but places on the rows and transitions on the cons. We construct a matrix that describes the input tokens that are consumed for a given transition and a matrix for those that have been produced:

\subsubsection{Consumption Matrix}
\[
W^{-} =
\begin{bmatrix}
    I_{1}(p_{1}) & I_{2}(p_{1}) & ... & I_{k}(p_{1}) \\
    I_{1}(p_{2}) & I_{2}(p_{2}) & ... & I_{k}(p_{2}) \\
    \vdots & \vdots & ... & \vdots \\
    I_{1}(p_{n}) & I_{2}(p_{n}) & ... & I_{k}(p_{n})
\end{bmatrix}
\]

\subsubsection{Production Matrix}
\[
W^{+} =
\begin{bmatrix}
    O_{1}(p_{1}) & O_{2}(p_{1}) & ... & O_{k}(p_{1}) \\
    O_{1}(p_{2}) & O_{2}(p_{2}) & ... & O_{k}(p_{2}) \\
    \vdots & \vdots & ... & \vdots \\
    O_{1}(p_{n}) & O_{2}(p_{n}) & ... & O_{k}(p_{n})
\end{bmatrix}
\]

\subsubsection{Incidence Matrix}
We then define the \textbf{incidence matrix} W that represent the global effect of every transition:

\begin{equation*}
    W = W^{+} - W^{-}
\end{equation*}

\section{Computing place invariants}
We have said that place invariants represent mass conservation, so all the reachable markings will preserve that constraint. Reachable markings are obtained by firing transition. Then every transition that fires has to preserve the overall weight of the matrix.\par
Then the weight of the tokens that are produced has to be the same as the weight of the tokens that are consumed in order to preserve the overall weight.\par
So I express invariant by focusing on individual transitions:\par
$\forall$ transition $t = <I, O>$ we should have:
\begin{equation*}
    \sum_{p \in P} I(p) \times i(p) = \sum_{p \in P}O(p) \times i(p)
\end{equation*}

which is the same as putting:
\begin{equation*}
    \sum_{p \in P} (O(p) - I(p)) \times i(p) = 0
\end{equation*}

Now considering the incidence matrix $W$ we have defined previously, any solution $i$ is a place invariant if satisfy the following scalar product:
\begin{equation*}
    i \times W = 0
\end{equation*}

\section{Reachability Tree Revised}
Another approach to overapproximation is based upon the Karp and Miller Tree. In general the reachability tree that describe the system behavior can be infinite. The approach here is to construct a "smarter" reachability tree that approximate the real reachability tree in a finite way. The approximation consist in replacing infinite paths with a special node.

\subsection{Monotonicity}
The construction of our overapproximated tree is based on a property of monotonicity present in the markings of Petri Nets. Petri Nets induce strongly monotonic transition systems. So what I do is condense markings in which the tokens remains the same.

\subsubsection{Example}
Given the following path of our tree where each elements reprents a marking and each row a transition:
\begin{equation*}
    <1, 0, 0, 0> \rightarrow <0, 0, 1, 0> \rightarrow <0, 0, 0, 1> \rightarrow <1, 0, 1, 1>
\end{equation*}

From the first three marking we can deduce that $p_{1}, p_{3}$ and $p_{4}$ are unbounded. Thus we can rewrite the last marking as:

\begin{equation*}
    <\omega, 0, \omega, \omega>
\end{equation*}

\textbf{Note:} $\omega$ must be regarded as: "any numer of tokens".

\subsubsection{Algorithm}
This is how we compute the successors of node n:

%insert image form slide 14 pag 109/147

\subsection{Stopping a branch}
Another small aspect to consider during the construction of the revised tree. On the current marking if I have to generate all the possible next markings considering all the possible transition that can be fired, if by firing a transition I generate a marking which is $\leq$ than a previous marking then I can cancel it, avoiding to continuing generating the marking there.

%insert image from lesson 14 pag 110/147i

\section{Properties of Karp and Miller Tree}

\subsection{Theorem 1}
A Karp and Miller tree is always finite
\subsubsection{Idea of the proof}
If the net is not bounded, int is because of some infinite increasing sequence of marking. Such sequences are detected in a finite amount of time by adding $\omega$ in the unbounded places.

\subsection{Theorem 2}
A net is bounded iff there is no node containing an $\omega$ in its Karp and Miller tree.

\subsection{Theorem 3}
A place $p$ is unbounded iff there exists a node labeled by $m$ in the Karp and Miller tree s.t. $m(p) = \omega$

\subsection{Theorem 4}
Transition $t$ is semi-live iff there exists a node labeled by $m$ in the Karp and Miller tree s.t. $t$ can fire in m.

\section{Coverability set}
First of all let's state the following properties:
\begin{itemize}
    \item The set of reachable markings \textbf{might be infinite}.
    \item The set of labels of the nodes of the Karp and Miller tree \textbf{are always finite}
\end{itemize}

We define the Coverability set as a set of markings related to the Karp and Miller tree that will be an overapproximation on the set of reachable markings.

\subsection{Downward-closure}
Let us assume that $\forall i \in \mathbb{N} : i < \omega$.
Let $m$ be a marking (possibly containing $\omega$), then its downward-closure is the set:
\begin{equation*}
    \downarrow m = \{m^{'} | m^{'} \preceq m\}
\end{equation*}

Given the set of markings $S = \{m_{1}, m_{2}, ..., m_{k}\}$ then:
\begin{equation*}
    \downarrow S = \downarrow m_{1} \cup \downarrow m_{2} \cup ... \cup \downarrow m_{k}
\end{equation*}

So the downward-closure operation can be used to construct a set of markings define as all the ones that are smaller than a given one.

\subsection{Theorems}
Let us state the following theorems:
\begin{itemize}
    \item $\downarrow Cover(N) = \downarrow Reach(N)$
    \item $Reach(N) \subseteq \downarrow Cover(N)$
\end{itemize}
hence $\downarrow Cover(N)$ is a finite over-approximation of $Reach(N)$.\par
This means that $Cover(N)$ is another overapproximation of the set of reachable markings. If a marking is not in $\downarrow Cover(N)$, \textbf{then it is not reachable} (sufficient condition).

\section{Another approach to reachability}
In Petri Nets reachability could be achieved not requiring a \textbf{exactly that amount} of tokens to reach the marking, we can reach a marking by having \textbf{at least} that number of tokens. \par
The problem thus becomes the coverability problem:
\begin{center}
    Does there exist a reachable marking which is larger than some marking b?
\end{center}

So given a marking $b$, to prove that it is reachable I start from the initial marking $m_{0}$ and apply the reachability graph and as it grows I check if it includes some marking $m^{'}$ which is greater or equal that $b$, in which case I can say that I can cover $b$.

\subsection{Formal definition}
There are two alternative definitions:
\begin{itemize}
    \item Is there a reachable marking $m$ s.t. $m \succeq b$?
    \item Does $Reach(N) \cap \{m | m \succeq b\} \neq \phi$ 
\end{itemize}

\subsection{First solution of the problem}
We use the coverability set since it is an over-approximation of the reachable states. The coverability set can be easily computed by using the Karp and Miller algorithm.\par
\subsubsection{Case where the marking is not reachable}
Consider:
\begin{equation*}
    \downarrow Cover(n) \cap U = \emptyset
\end{equation*}

implies

\begin{equation*}
    Reach(N) \cap U = \emptyset
\end{equation*}

\subsubsection{Case where the marking is reachable}
There is $m \in \downarrow Cover(N) \cup U$, hence there is $m^{'} \succeq m$ which is in $Reach(N)$. However, any $m^{'} \succeq m$ is also in U. Thus there is a $m^{'}$ both in $Reach(N)$ and $U$.

Note: $m^{'} > m$ since $\downarrow Cover = \downarrow Reach$.\par

So now we have the following implication:

\begin{equation*}
    Reach(N) \cap U = \emptyset
\end{equation*}

implies

\begin{equation*}
    \downarrow Cover(N) \cap U = \emptyset
\end{equation*}

Thus we can say:
\subsubsection{Theorem}

\begin{equation*}
    Reach(N) \cap U = \emptyset \iff \downarrow Cover(N) \cap U = \emptyset
\end{equation*}

Which means that in order to check whether a marking in U is reachable, we can use the Karp and Miller approach to compute the $\downarrow Cover(N)$ set and check whether it has a non-empty intersection with U.

\section{Remark regard our overapproximation methods}
Note how all of the method we listed does not limit the state space, like putting an artificial bounds to the number of tokens that can be present. In Prims we require limitation on the state space because the language it uses is more richer than the one from Petri Nets.